---
title: "Mousetrap validation"
author: "Pascal J. Kieslich & Felix Henninger"
output: html_document
---

# Validation settings
* OpenSesame: version 3.1.6 with legacy backend
* Mousetrap-os plugin: version 1.2.1
* Computer: Windows 7 Professional, Intel Pentium Dual-Core 3 GHz, 4 GB RAM
* External hardware (Henninger, 2017) used to generate predetermined movement patterns
* Cursor position updated at the logging resolution (10 ms) 
* Two simulations with 1000 trials each


# General preparation
```{r,message=FALSE}
# Load libraries
library(readbulk)
library(mousetrap)
library(dplyr)
library(ggplot2)

# Set custom ggplot2 theme
theme_set(theme_classic()+ 
  theme(
    axis.line = element_line(colour = "black"),
    axis.ticks = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    panel.border = element_rect(colour = "black", fill=NA),
    strip.background = element_rect(colour = NA)
  ))

options(width=120)
```


# Plot trajectories
```{r}
raw_data <- read_opensesame("validation_data")
mt_data <- mt_import_mousetrap(raw_data)
mt_data <- mt_remap_symmetric(mt_data)
mt_data$data$Condition <- factor(mt_data$data$subject_nr,levels=c(1,2), labels=c("Diagonal","Triangular"))
mt_plot(mt_data,facet_col = "Condition")+
  xlab("x coordinate (px)") + ylab("y coordinate (px)")
# ggsave("FigureA1.pdf", width = 16, height=9, unit="cm")
# ggsave("FigureA1.png", width = 16, height=9, unit="cm", dpi=600)
```


# Validation 1: Diagonal path
* Start click (0,-400) followed by 110 ms pause
* Every 10 ms cursor moves both one pixel up and one left for 800 px, i.e., for 8000 ms in total
* Cursor pauses at end position (-800,400) for 100 ms and then clicks

## Read and preprocess data
```{r}
raw_data <- read_opensesame("validation_data",extension = "diagonal.csv.gz")

mt_data <- mt_import_mousetrap(raw_data)
mt_data <- mt_remap_symmetric(mt_data,remap_xpos = "no")
mt_data <- mt_measures(mt_data)
mt_data <- mt_derivatives(mt_data, return_delta_time = TRUE, dimensions = "xpos", prefix = "xpos_")
mt_data <- mt_derivatives(mt_data, return_delta_time = TRUE, dimensions = "ypos", prefix = "ypos_")

mt_data_no_reset <- mt_import_mousetrap(raw_data,reset_timestamps = FALSE)
```

## Temporal analyses
```{r}
mt_data$data <- mt_data$data %>% 
  mutate(
    time_start_click = time_get_start_click+response_time_get_start_click,
    delta_click_stimulus = time_present_stimulus-time_start_click,
    delta_stimulus_tracking = mt_data_no_reset$trajectories[,1,"timestamps"]-time_present_stimulus,
    delta_click_tracking = mt_data_no_reset$trajectories[,1,"timestamps"]-time_start_click
  )

summary(select(mt_data$data,starts_with("delta"),response_time),digits = 8)
mt_data$data %>% select(starts_with("delta"),response_time) %>% summarise_all(c("sd"))

# mousetrap-os response_time matches mt_measures RT
table(mt_data$data$response_time==mt_data$measures$RT)
```

## Logging resolution
```{r}
mt_check_resolution(mt_data,desired = 10)

# Percent of lags that lag by 1 ms only
147/148
```

## Distances between subsequently recorded positions
```{r}
# Frequency of distances converge across both x and y coordinates
table(as.numeric(mt_data$trajectories[,-1,"xpos_dist"]),as.numeric(mt_data$trajectories[,-1,"ypos_dist"]))

# Frequency of distances
table(as.numeric(mt_data$trajectories[,-1,"xpos_dist"]))

# Percent of distances
round(table(as.numeric(mt_data$trajectories[,-1,"xpos_dist"]))/sum(!is.na(mt_data$trajectories[,-1,"xpos_dist"])),6)
```

## Comparison of expected and observed position
```{r}
# Read in raw data from hardware that generated mouse movements
mouse_coordinates <- read.csv("mouse_diagonal.csv",sep=",",col.names = c("xpos","ypos","click"))

# Create data frame with expected position for each timestamp
expected <- mouse_coordinates[rep(seq(which(mouse_coordinates$click==1)[1],which(mouse_coordinates$click==1)[2]),each=10),]
expected$ypos <- (-expected$ypos)

# Set constant for delay between start click and tracking onset
delta_tracking_onset <- 7

# Determine expected position taking delay between start click and tracking onset into account
mt_data <- mt_add_variables(mt_data,use="trajectories",variables = c("xpos_expected","ypos_expected"))
for (i in rownames(mt_data$trajectories)){
  mt_data$trajectories[i,,"xpos_expected"] <- expected[mt_data$trajectories[i,,"timestamps"]+delta_tracking_onset,"xpos"]
  mt_data$trajectories[i,,"ypos_expected"] <- expected[mt_data$trajectories[i,,"timestamps"]+delta_tracking_onset,"ypos"]
}

# Correlation between observed and expected position
cor_xpos <- cor(as.vector(mt_data$trajectories[,,"xpos"]),as.vector(mt_data$trajectories[,,"xpos_expected"]),use="complete.obs")
cor_ypos <- cor(as.vector(mt_data$trajectories[,,"ypos"]),as.vector(mt_data$trajectories[,,"ypos_expected"]),use="complete.obs")
print(cor_xpos,digits = 15)
print(cor_ypos,digits = 15)

# Compute difference between expected and observed position
mt_data <- mt_add_variables(mt_data,use="trajectories",variables=list(
  xpos_diff = mt_data$trajectories[,,"xpos_expected"]-mt_data$trajectories[,,"xpos"],
  ypos_diff = mt_data$trajectories[,,"ypos_expected"]-mt_data$trajectories[,,"ypos"]
))

# Frequency of differences between observed and expected positions across both dimensions
# --> differences converge across both dimentions
table(mt_data$trajectories[,,"xpos_diff"],mt_data$trajectories[,,"ypos_diff"])

# Percent of differences
round(table(mt_data$trajectories[,,"xpos_diff"])/sum(!is.na(mt_data$trajectories[,,"xpos_diff"])),6)
```

## Mouse-tracking indices
```{r}
summary(select(mt_data$measures,MAD,AUC,AD),digits = 8)
mt_data$measures %>% select(MAD,AUC,AD) %>% summarise_all(c("sd"))
```


# Validation 2: Triangular path
* Start click (0,-400) followed by 110 ms pause
* Every 10 ms cursor moves one pixel up for the first 800 px and then left for 800 px, i.e., for 16000 ms in total
* Cursor pauses at end position (-800,400) for 100 ms and then clicks

## Read and preprocess data
```{r}
raw_data <- read_opensesame("validation_data",extension = "triangular.csv.gz")

mt_data <- mt_import_mousetrap(raw_data)
mt_data <- mt_remap_symmetric(mt_data,remap_xpos = "no")
mt_data <- mt_measures(mt_data)
mt_data <- mt_derivatives(mt_data, return_delta_time = TRUE, dimensions = "xpos", prefix = "xpos_")
mt_data <- mt_derivatives(mt_data, return_delta_time = TRUE, dimensions = "ypos", prefix = "ypos_")

mt_data_no_reset <- mt_import_mousetrap(raw_data,reset_timestamps = FALSE)
```

## Temporal analyses
```{r}
mt_data$data <- mt_data$data %>% 
  mutate(
    time_start_click = time_get_start_click+response_time_get_start_click,
    delta_click_stimulus = time_present_stimulus-time_start_click,
    delta_stimulus_tracking = mt_data_no_reset$trajectories[,1,"timestamps"]-time_present_stimulus,
    delta_click_tracking = mt_data_no_reset$trajectories[,1,"timestamps"]-time_start_click
  )

summary(select(mt_data$data,starts_with("delta"),response_time),digits = 8)
mt_data$data %>% select(starts_with("delta"),response_time) %>% summarise_all(c("sd"))

# validate mousetrap-os response_time with mt_measures RT
table(mt_data$data$response_time==mt_data$measures$RT)
```

## Logging resolution
```{r}
mt_check_resolution(mt_data,desired = 10)

# Percent of lags that lag by 1 ms only
284/301
```

## Distances between subsequently recorded positions
```{r}
# Frequency of distances across both x and y coordinates
table(as.numeric(mt_data$trajectories[,-1,"xpos_dist"]),as.numeric(mt_data$trajectories[,-1,"ypos_dist"]))

# Frequency of distances for x coordinates
table(as.numeric(mt_data$trajectories[,-1,"xpos_dist"]))

# Percent of distances for x coordinates
round(table(as.numeric(mt_data$trajectories[,-1,"xpos_dist"]))/sum(!is.na(mt_data$trajectories[,-1,"ypos_dist"])),6)

# Frequency of distances for y coordinates
table(as.numeric(mt_data$trajectories[,-1,"ypos_dist"]))

# Percent of distances for y coordinates
round(table(as.numeric(mt_data$trajectories[,-1,"ypos_dist"]))/sum(!is.na(mt_data$trajectories[,-1,"ypos_dist"])),6)
```

## Comparison of expected and observed position
```{r}
# Read in raw data from hardware that generated mouse movements
mouse_coordinates <- read.csv("mouse_triangular.csv",sep=",",col.names = c("xpos","ypos","click"))

# Create data frame with expected position for each timestamp
expected <- mouse_coordinates[rep(seq(which(mouse_coordinates$click==1)[1],which(mouse_coordinates$click==1)[2]),each=10),]
expected$ypos <- (-expected$ypos)

# Set constant for delay between start click and tracking onset
delta_tracking_onset <- 7

# Determine expected position taking delay between start click and tracking onset into account
mt_data <- mt_add_variables(mt_data,use="trajectories",variables = c("xpos_expected","ypos_expected"))
for (i in rownames(mt_data$trajectories)){
  mt_data$trajectories[i,,"xpos_expected"] <- expected[mt_data$trajectories[i,,"timestamps"]+delta_tracking_onset,"xpos"]
  mt_data$trajectories[i,,"ypos_expected"] <- expected[mt_data$trajectories[i,,"timestamps"]+delta_tracking_onset,"ypos"]
}

# Correlation between observed and expected position
cor_xpos <- cor(as.vector(mt_data$trajectories[,,"xpos"]),as.vector(mt_data$trajectories[,,"xpos_expected"]),use="complete.obs")
cor_ypos <- cor(as.vector(mt_data$trajectories[,,"ypos"]),as.vector(mt_data$trajectories[,,"ypos_expected"]),use="complete.obs")
print(cor_xpos,digits = 15)
print(cor_ypos,digits = 15)

# Compute difference between expected and observed position
mt_data <- mt_add_variables(mt_data,use="trajectories",variables=list(
  xpos_diff = mt_data$trajectories[,,"xpos_expected"]-mt_data$trajectories[,,"xpos"],
  ypos_diff = mt_data$trajectories[,,"ypos_expected"]-mt_data$trajectories[,,"ypos"]
))

# Frequency of differences between observed and expected positions across both dimensions
table(mt_data$trajectories[,,"xpos_diff"],mt_data$trajectories[,,"ypos_diff"])

# Percent of differences for xpos
round(table(mt_data$trajectories[,,"xpos_diff"])/sum(!is.na(mt_data$trajectories[,,"xpos_diff"])),6)

# Percent of differences for ypos
round(table(mt_data$trajectories[,,"ypos_diff"])/sum(!is.na(mt_data$trajectories[,,"ypos_diff"])),6)
```

## Mouse-tracking measures
```{r}
# Descriptives
summary(select(mt_data$measures,MAD,AUC,AD),digits = 8)
mt_data$measures %>% select(MAD,AUC,AD) %>% summarise_all(c("sd"))

# Expected MAD
.5*sqrt(800^2+800^2)

# Expected AUC
.5*800^2

# Expected AD
mean(c(seq(0,800,1),seq(799,0,-1))/sqrt(2))*1601/1622
```


